Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-04-21 doc/invt.tex
add histogram flatness == error

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
The above two introductory sections help define our notations,
and an expert reader may wish to skip them.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses metadynamics
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix},
and present a general procedure of computing the optimal schedule
in Sec. \ref{sec:procedure}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time or extensive sampling limit
in Sec. \ref{sec:cmpschemes}.

@@

In terms of how one adjusts the bias potential,
there are roughly two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(estimated a priori)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, *lee1993}
and simulated tempering\cite{
marinari1992, *lyubartsev1992}.
%with $i$ being the index of energy
%and that of temperature, respectively.
%
After a sufficiently long run, the bias potential, $V_i$,
can often be improved
with the normalized histogram, $H_i$, from the simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_V}
allows us to correct $V_i$ as
%
\begin{equation}
  V^\mathrm{corr}_i
  =
  V_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
%However, for an unknown system,
%it is often difficult
%to guess a sufficiently accurate
%$V_i$ with which to start.
%%
%Thus,
%it can take many iterative rounds of simulations
%to achieve the desired flat distribution.
%%
%%Another difficulty of this approach is that
%The effectiveness of this approach
%depends on the precision of the histogram, $H_i$,
%which in turn depends on the simulation length
%that must be decided in advance.
However, this formula often requires a long
simulation period to accumulate a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

@@

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution, $\pi_i(t)$,
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
In this way,
the variable, $x_i(t)$,
represents the actual deviation of the bias potential, $V_i(t)$,
and the total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_sum}
\end{equation}

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
%
where the modified histogram fluctuation
%
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$,
i.e. we may write
$$
  \langle \xi_i(t) \, \xi_i(t') \rangle = G_i \, \delta(t - t'),
$$
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

The total error
at the end of the simulation, $t = T$,
as defined in Eq. \eqref{eq:error_sum},
now depends implicitly on the function $\alpha(t)$
via $x_i(T)$.
%
Our aim is to find the $\alpha(t)$
that minimizes this error.
%%
%It suffices to consider
%a one-variable problem that
%minimizes each individual
%$\left\langle x_i^2(T) \right\rangle$.

@@

We recall that
the formal solution of Eq. \eqref{eq:dxdt_WL} is
%
\begin{equation}
x_i(T)
=
x_i(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi_i(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}
%
At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_sum}
is given by
%
%\begin{align}
%  &\left\langle x_i^2(T) \right\rangle
%  =
%  \langle x_i^2(0) \rangle \, e^{-2 \, q(T)}
%  +
%  G_i \int_0^T
%    {\dot u}^2\bigl( q(t) \bigr) \,
%    \, dt.
%  \label{eq:x2t_average}
%\end{align}
%
\begin{align}
  E(T)
  =
  E(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot u}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where
$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.

@@

Since the above optimization is performed
under a fixed $q(T)$,
it can be improved
by choosing a proper value of $q(T)$.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using the optimal schedule in Eq. \eqref{eq:ET_average},
we find the final total error to be
\begin{equation}
  E(T)
  =
  \frac{ E(0) \, t_0^2 + \Gamma \, T}
       { (t_0 + T)^2 }
  ,
  \notag
  %\label{eq:error_singlebin}
\end{equation}
%
which is minimized if
%
\begin{equation}
t_0 = \frac{ \Gamma }
           { E(0) }
           ,
  \label{eq:t0_singlebin}
\end{equation}
%
and at this point
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}

@@

Diagonalizing the matrix, $\mathbf w$,
using the eigenvectors defined in Eqs.
\eqref{eq:eig_w}-\eqref{eq:eig_orthonormal_rows}
yields a set of decoupled equations
%
\begin{equation}
  \dot y_k(t)
  =
  -\alpha(t) \, \lambda_k \, [y_k(t) - \theta_k(t)],
\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_k(t)
  &\equiv
  \sum_{j=1}^n
    \phi_{kj} \, x_j(t)
  ,
  \label{eq:y_def}
\end{align}
and
\begin{align}
  \theta_k(t)
  &\equiv
  \sum_{j=1}^n
    \phi_{kj} \, \xi_j(t)
  ,
  \label{eq:theta_def}
\end{align}
%
are the deviation and noise (histogram fluctuation)
of mode $k$, respectively.

@@

The error function defined in Eq. \eqref{eq:error_sum}
can be rewritten in terms of the eigenmodes of $\mathbf w$,
%
\begin{align}
  \Err
  \equiv
  \sum_{i = 1}^n \rho_i \,
    \left\langle x_i^2 \right\rangle
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \,
    \left\langle x_i \, x_j \right\rangle
  =
  \sum_{k = 1}^n
    \left\langle
      y_k^2
    \right\rangle
  ,
\notag
\end{align}
%
where we have used Eq. \eqref{eq:eig_orthonormal_cols}.
%
From Eq. \eqref{eq:eigenmode1},
the first eigenmode,
$y_1 = \sum_{j=1}^n \rho_j \, x_j$,
vanishes identically by Eq. \eqref{eq:rhox_sum}.
%
Thus, we can start the last sum %in Eq. \eqref{eq:y2_sum}
from $k = 2$,
%
\begin{equation}
  \Err(T)
  =
  \sum_{k = 2}^n
    \left\langle
    y_k^2(T)
    \right\rangle
  .
\label{eq:y2_sum}
\end{equation}

@@

\subsection{\label{sec:mbin_opta}
Optimal schedule for multiple-bin updating schemes}

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 2 }^n
  %  \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(Q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
\label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,
and we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$
%
This serves
as a generalization of Eq. \eqref{eq:dudt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(Q)
    \;
    d Q
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    M( Q )
    \;
    d Q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

@@

\subsection{\label{sec:band-matrix}
Homogeneous updating schemes}

@@

\subsection{\label{sec:procedure}
Procedure for computing the optimal schedule and error
}

@@

By including the residual error,
we determine the optimal $t_0$
that minimizes the total error.
%
According to Eq. \eqref{eq:half_alpha0}, we have
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  .
\label{eq:t0_sinc}
\end{equation}
%
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:y2_eql}.
%
From Eqs. \eqref{eq:error_tot},
\eqref{eq:error_res1},
\eqref{eq:tk0},
and
\eqref{eq:error_asym_sinc},
we get
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
which serves as a generalization of
the result from the single-bin scheme,
Eq. \eqref{eq:Emin_singlebin}
[one can show that $\sum_{k=2}^n \Gamma_k = \sum_{i=1}^n \rho_i \, G_i = \Gamma$,
following the steps similar to those leading to Eq. \eqref{eq:y2_sum}].
%
\note{The derivation.
From Eq. \eqref{eq:eig_orthonormal_cols}, we have
$\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle
=\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle$,
so
\begin{align*}
\sum_{i=1}^n \rho_i G_i
&=\sum_{t=-\infty}^\infty\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle \\
&=\sum_{t=-\infty}^\infty\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle
=\sum_{k=2}^n \Gamma_k.
\end{align*}
}
%
This error is roughly proportional to the inverse simulation length,
as one would expect from an equilibrium FDS simulation.
%In Appendix \ref{sec:equilerr},
%we show that the above error tends
%to the error of the ideal equilibrium FDS simulation
%in the asymptotic limit.

@@

>>>>

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
The above two introductory sections help define our notations,
and an expert reader may wish to skip them.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian scheme used in metadynamics
in Sec. \ref{sec:multiple-bin},
and present a general procedure of computing the optimal schedule
in Sec. \ref{sec:procedure}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time or extensive sampling limit
in Sec. \ref{sec:cmpschemes}.

@@

In terms of how one adjusts the bias potential,
there are roughly two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(estimated a priori)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, *lee1993}
and simulated tempering\cite{
marinari1992, *lyubartsev1992}.
%with $i$ being the index of energy
%and that of temperature, respectively.
%
After a sufficiently long run, the bias potential, $V_i$,
can often be improved
with the normalized histogram, $H_i$,
accumulated from the simulation as
%
%Using $H_i$ as an estimate of $\pi_i$ in
%Eq. \eqref{eq:pi_p_V}
%allows us to correct $V_i$ as
%
\begin{equation}
  V^\mathrm{corr}_i
  =
  V_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
%However, for an unknown system,
%it is often difficult
%to guess a sufficiently accurate
%$V_i$ with which to start.
%%
%Thus,
%it can take many iterative rounds of simulations
%to achieve the desired flat distribution.
%%
%%Another difficulty of this approach is that
%The effectiveness of this approach
%depends on the precision of the histogram, $H_i$,
%which in turn depends on the simulation length
%that must be decided in advance.
However, this formula often requires a long
accumulation period for a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

@@

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution, $\pi_i(t)$,
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
In this way,
the variable, $x_i(t)$,
represents the actual deviation of the bias potential, $V_i(t)$,
and the total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
%
where the shifted histogram fluctuation
%
\begin{equation}
  \xi_i(t)
  \equiv
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$,
i.e. we may write
\begin{equation}
  \langle \xi_i(t) \, \xi_i(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

The total error
at the end of the simulation, $t = T$,
as defined in Eq. \eqref{eq:error_def},
now depends implicitly on the function $\alpha(t)$
via $x_i(T)$.
%
Our aim is to find the $\alpha(t)$
that minimizes this error.
%%
%It suffices to consider
%a one-variable problem that
%minimizes each individual
%$\left\langle x_i^2(T) \right\rangle$.

@@

We recall that
the formal solution of Eq. \eqref{eq:dxdt_WL} is
%
\begin{equation}
x_i(T)
=
x_i(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi_i(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}
%
At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
%\begin{align}
%  &\left\langle x_i^2(T) \right\rangle
%  =
%  \langle x_i^2(0) \rangle \, e^{-2 \, q(T)}
%  +
%  G_i \int_0^T
%    {\dot u}^2\bigl( q(t) \bigr) \,
%    \, dt.
%  \label{eq:x2t_average}
%\end{align}
%
\begin{align}
  E(T)
  =
  E(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot u}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where
\begin{equation}
  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
  \label{eq:Gamma_def}
\end{equation}

@@

Since the above optimization is performed
under a fixed $q(T)$,
it can be improved
by choosing a proper value of $q(T)$.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using the optimal schedule in Eq. \eqref{eq:ET_average},
we find the final total error to be
%
$E(T) = [ E(0) \, t_0^2 + \Gamma \, T ]/(t_0 + T)^2$,
%
which is minimized if
%
\begin{equation}
t_0 = \frac{ \Gamma }
           { E(0) }
           ,
  \label{eq:t0_singlebin}
\end{equation}
%
and at this point, the error is given by
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}
%
This error is roughly equal to the overall histogram flatness
and thus it can be conveniently estimated\cite{zhou2005, zhou2008}
(see also Appendix \ref{sec:flatness}).

@@

Diagonalizing the matrix, $\mathbf w$,
using the eigenvectors defined in Eqs.
\eqref{eq:eig_w}-\eqref{eq:eig_orthonormal_rows}
yields a set of decoupled equations
%
\begin{equation}
  \dot y_k(t)
  =
  -\alpha(t) \, \lambda_k \, [y_k(t) - \theta_k(t)],
  \notag
  %\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_k(t)
  &\equiv
  \sum_{j=1}^n
    \phi_{kj} \, x_j(t)
  ,
  %\label{eq:y_def}
\end{align}
and
\begin{align}
  \theta_k(t)
  &\equiv
  \sum_{j=1}^n
    \phi_{kj} \, \xi_j(t)
  ,
  \label{eq:theta_def}
\end{align}
%
are the deviation and noise (histogram fluctuation)
of mode $k$, respectively.

@@

The error function defined in Eq. \eqref{eq:error_def}
can be rewritten in terms of the eigenmodes of $\mathbf w$,
%
\begin{align}
  \Err
  \equiv
  \sum_{i = 1}^n \rho_i \,
    \left\langle x_i^2 \right\rangle
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \,
    \left\langle x_i \, x_j \right\rangle
  =
  \sum_{k = 1}^n
    \left\langle
      y_k^2
    \right\rangle
  ,
\notag
\end{align}
%
where we have used Eq. \eqref{eq:eig_orthonormal_cols}.
%
From Eq. \eqref{eq:eigenmode1},
the first eigenmode,
$y_1 = \sum_{j=1}^n \rho_j \, x_j$,
vanishes identically by Eq. \eqref{eq:rhox_sum}.
%
Thus, we can start the last sum %in Eq. \eqref{eq:y2_sum}
from $k = 2$,
%
\begin{equation}
  \Err(T)
  =
  \sum_{k = 2}^n
    \left\langle
    y_k^2(T)
    \right\rangle
  .
\label{eq:y2_sum}
\end{equation}

@@

%\subsection{\label{sec:mbin_opta}
%Optimal schedule for multiple-bin updating schemes}

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 2 }^n
  %  \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(Q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
\label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,
and we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$.
%
This serves
as a generalization of Eq. \eqref{eq:dudt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(Q)
    \;
    d Q
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    M( Q )
    \;
    d Q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

@@

\subsubsection{\label{sec:band-matrix}
Homogeneous updating schemes}

@@

\subsubsection{\label{sec:procedure}
Procedure for computing the optimal schedule and error
}

@@

By including the residual error,
we determine the optimal $t_0$
that minimizes the total error.
%
According to Eq. \eqref{eq:half_alpha0}, we have
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  .
\label{eq:t0_sinc}
\end{equation}
%
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:y2_eql}.
%
From Eqs. \eqref{eq:error_tot},
\eqref{eq:error_res1},
\eqref{eq:tk0},
and
\eqref{eq:error_asym_sinc},
we get
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
In particular, for $K = n$, we have
\begin{equation}
  \sum_{k=2}^n \Gamma_k
  = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  ,
  \notag
  %\label{eq:Gamma_sum}
\end{equation}
which can be shown by following the steps
similar to those leading to Eq. \eqref{eq:y2_sum},
and the above error recovers the single-bin result,
Eq. \eqref{eq:Emin_singlebin}.
%
%
%
\note{The derivation.
From Eq. \eqref{eq:eig_orthonormal_cols}, we have
$\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle
=\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle$,
so
\begin{align*}
\sum_{i=1}^n \rho_i G_i
&=\sum_{t=-\infty}^\infty\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle \\
&=\sum_{t=-\infty}^\infty\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle
=\sum_{k=2}^n \Gamma_k.
\end{align*}
}
%
This error is roughly proportional to the inverse simulation length,
as one would expect from an equilibrium FDS simulation.
%In Appendix \ref{sec:equilerr},
%we show that the above error tends
%to the error of the ideal equilibrium FDS simulation
%in the asymptotic limit.

@@

\section{\label{sec:flatness}
Histogram flatness}
